---
title: "Is soybean use driving global deforestation?"
author: by Team 5 Look Alive
output: github_document
---

```{r load-packages, echo = FALSE, message = F, warning = F}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(ggdark)
library(scales)
library(viridis)
library(gghighlight)
knitr::opts_chunk$set(echo = F)
```


## Introduction

Our data comes originally from the UN Food and Agriculture Organization’s Forest Resources Assessment. "Since year-to-year changes in forest cover can be volatile, the UN FAO provide this annual data averaged over five-year periods." It measures the area of forests over time in countries around the world. It also holds more detailed information about the loss of Brazilian forest over time. The data further holds data about soybean production and vegetable oil production by year and country.

The dataset has 5 elements: `forest`, `forest_area`, `brazil_loss`, `soybean_use`, and `vegetable_oil`. The set `forest` contains the change every 5 years in for forest area in conversion. `forest_area` contains the change in global forest area as a percent of global forest area. `brazil_loss` contains the loss of Brazilian forest due to specific types of deforestation, such as `fire` and `small_scale_clearing`. `soybean_use` contains the soybean production and use by year and by country. Finally, the set `vegetable_oil` contains vegetable oil production by crop type and year.

## What countries are seeing the most deforestation within their borders, and what is causing this deforestation?

### Introduction

These ideas stuck us as the most essential elements of deforestation, and gaining an understanding of the complexities of deforestation depends on becoming familiar with these concepts.

### Approach

In this question, we planned on exploring three things: the who, the what, and the why of deforestation. First, we chose to use a line plot to examine trends over time in deforestation. This would allow us to get a rough understanding of how many countries are facing the most deforestation and examine the most deforested country’s data, using a cumulative line plot of deforestation and a bar plot to compare across years. From here, we could then examine the “what”. To do this, we decided to compare the proportions with an area plot. This gave us the opportunity to showcase which causes are the most impactful in Brazil’s deforestation. Finally, to understand the “why” of deforestation, we selected to examine one of the leading causes of deforestation in Brazil. Using line plots again allowed us to highlight which regions and countries are most influential in driving soybean production globally.

Putting all of these factors together gives us a basic understanding of global deforestation.

### Analysis

#### Figure 1

```{r cumulative_brazil_only,  message = F, fig.alt="Brazil lost 30m hectares of forest from 2000 to 2013."}

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))

# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )

forest <- read.csv("data/forest.csv")
forest_area <- read.csv("data/forest_area.csv")
brazil_loss <- read.csv("data/brazil_loss.csv")
vegetable_oil <- read.csv("data/vegetable_oil.csv")
soybean_use <- read.csv("data/soybean_use.csv")

brazil_loss$total_brazil_forest_loss_hectares = rowSums(brazil_loss[,5:15])

annotation <- data.frame(
   x = c(2010),
   y = c(-7300000),
   label = c("Brazil has lost")
)
annotation2 <- data.frame(
   x = c(2010),
   y = c(-10500000),
   label = c("over 30 million")
)
annotation3 <- data.frame(
   x = c(2010),
   y = c(-15700000),
   label = c("hectares of forest\nsince 2000")
)

brazil_loss %>% select(year, total_brazil_forest_loss_hectares) %>%
  add_row(year = 2000, total_brazil_forest_loss_hectares = 0, .before = 1) %>%
  mutate(cum_sum = -(cumsum(total_brazil_forest_loss_hectares))) %>%
  ggplot() +
  geom_line(aes(year, cum_sum), size = 2, color = "#009739") +
  geom_bar(aes(year, -total_brazil_forest_loss_hectares), stat="identity") +
  geom_hline(aes(yintercept = 0), color = "red", size = 1.5) +
  labs(title = "Cumulative Brazilian forest loss", subtitle = "2000 to 2013", x = NULL, y = "Hectares") +
  scale_y_continuous(labels = label_number_si()) +
  scale_x_continuous() +
  geom_text(data=annotation, aes(x=x, y=y, label=label),
           color="black", 
           size=6, angle=0) +
  geom_text(data=annotation2, aes(x=x, y=y, label=label),
           color="red", 
           size=7, angle=0, fontface="bold") +
  geom_text(data=annotation3, aes(x=x, y=y, label=label),
           color="black", 
           size=6, angle=0) +
  scale_x_continuous(breaks = seq(from = 2000, to = 2013, by = 4)) +
  theme(panel.grid.minor.x = element_blank())
```

#### Figure 2

```{r drivers_of_brazil_deforstation,  message = F, fig.alt = "Pasture is the most common cause of deforestation in Brazil, "}
brazil_loss_no_total <- read.csv("data/brazil_loss.csv") %>% select(-(1:3))
brazil_loss_no_total <- brazil_loss_no_total %>%
  pivot_longer(!year,
               names_to = "cause",
               values_to = "forest_lost") %>%
  mutate(cause = case_when(
    cause == "pasture" ~ "Pasture",
    cause == "fire" ~ "Fire",
    cause == "commercial_crops" ~ "Commercial Crops",
    cause == "selective_logging" ~ "Logging",
    cause == "small_scale_clearing" ~ "Small-Scale Farming",
    T ~ "Other"
  ) %>% 
  factor(levels=c('Fire', "Logging", "Small-Scale Farming", "Commercial Crops", 'Pasture',"Other"))) %>%
  group_by(cause, year) %>%
  summarise(sum = sum(forest_lost), .groups = 'drop') %>%
  group_by(year, cause) %>%
  summarise(n = sum(sum)) %>%
  mutate(percentage = n / sum(n))

ggplot(brazil_loss_no_total, aes(year, percentage, fill=cause)) + 
    geom_area(alpha = 0.9) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#CC79A7", "red", "#0072B2", "#D55E00")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = NULL, title = "Causes of loss of forest in Brazil", fill = NULL,
       caption = "Fig. 2") +
  scale_x_continuous(breaks = seq(from = 2000, to = 2013, by = 4)) +
  theme(legend.position = "bottom")
```

#### Figure 3

```{r soybean_use_countries,  message=F, fig.alt = "China, US, Brazil, Argentina identified as countries that use the most soybean."}
soy_use_by_country <- soybean_use %>%
  filter(!is.na(code),
         entity != "World") %>%
  rowwise() %>%
  mutate(total = sum(human_food, animal_feed, processed, na.rm = T))

# https://stackoverflow.com/questions/49438953/selective-labeling-for-ggplot-lines
# https://cran.r-project.org/web/packages/gghighlight/vignettes/gghighlight.html

soy_use_by_country %>%
  ggplot(aes(x = year, y = total, color = entity)) +
  geom_line(show.legend = F, size = 2) +
  gghighlight(max(total), max_highlight = 4,
              unhighlighted_params = list(size = 1.2, colour =
                                            alpha("gray", 0.6))) +
  scale_y_continuous(labels = label_number_si()) +
  labs(y = "Tonnes", x = NULL, title = "Soybean use by country",
       caption = "Fig. 3") +
  # flagcolorcodes.com
  scale_color_manual(values = c("#6CACE4", "#009739", "#EE1C25", "#0A3161")) +
  scale_x_continuous(breaks = seq(from = 1960, to = 2013, by = 10))
```

### Discussion

Figure 1 portrays the amount of lost forest in Brazil in the span of 13 years, from 2000 to 2013. The loss is striking at over 30 million hectares (300,000 square km, for reference). The figure does show a concave trend, which shows that the loss of forest is not exponential, signaling that there may be hope in a continued flattening of this line of forest loss. It is interesting to consider what factors may have contributed to these trends within Brazil. Lawmaking and leadership will play a role in a country's development, meaning that it would be fascinating to gather this same data from Jair Bolsonaro's current reign as Brazil's president and compare it to the trends under Brazilian presidents who led during the time period of this plot, such as Lula de Silva, who served as president from 2003-2010 in Brazil (Wikipedia, Brazilian Presidents).

Figure 2 offers a view of the causes of Brazilian loss of forest. The most common reason for removal of forest in Brazil is to clear space for land to be used as pasture. Other big causes of loss of forest include fire, logging, small-scale farming, and commercial crops. Fires fluctuate wildly on the plot, demonstrating their lack of predictability and their destructive nature. Global warming and human activity may be factors that contributes to the increased danger of forest fires in the Amazon (Bush et. al. 2008). One important commercial crop is soybean, which we chose to analyze further, guided by our retrieval of data on soybean use.

Figure 3 gives a look into which countries are the largest consumers of soybean, which in turn increases demand for soybean and drives deforestation. China, the United States, Brazil, and Argentina were found to be the most prevalent soybean consumers. Argentina was the most surprising to us, but we found that soybean is a vital crop to Argentina, which incentivizes the use of soybean after its growth (croplife, Soybeans in Argentina).

## How does the proportions of soybean production used for human food versus for feeding animal change over time? Is this change in proportions related to global deforestation?

### Introduction

We want to examine the proportions of soybean production used for human food versus for feeding animal change over time and whether this change in proportions is related to global deforestation development over time. The datasets we will use for this question are `soybean_use` and `forest`.

### Approach

1. Start by grouping all entries in `soybean_use.csv` by years. For each year, add up the values for all entries under `human_food` and `animal_feed`.

2. Create new variables `prop` that represents the proportions of soybean production used for human food and for animal
feed in each year.

3. Use a stacked bar chart to plot the difference in proportions of soybean use between human food and animal feed via different fill colors. The x-axis represents `year` and the y-axis represents 'prop'.

4. Use a line graph to plot the negative net conversion of global forest area in 'forest.csv', and an area graph to comparatively represent the remaining forest area.


### Analysis

#### Figure 1

```{r soybean_use_share,  message = FALSE, fig.alt = "global soybean production share by animal feed is catching up to, if not exceeding, the share by human food between 1961 and 2013"}

data2 <-soybean_use %>%
  group_by(year) %>%
  summarise(total_human = sum(human_food, na.rm = T),
            total_animal = sum(animal_feed, na.rm = T)) %>%
  pivot_longer(!year, 
              names_to = "use",
              values_to = "amount") 

data2 <- data2 %>%
   group_by(year) %>%
   mutate(prop = amount / sum(amount))
 
ggplot(data2, aes(fill = use, y = prop, x = year)) + 
  dark_theme_light() +  
  geom_bar(position ="fill", stat ="identity") +
  
  scale_x_continuous (name = "Year",
                     breaks = seq(from = 1961, to = 2013, by = 10)) +
  scale_y_continuous(name = "Proportion",
                     labels = c("0%", "25%", "50%","75%","100%")) +
  scale_fill_manual(values = c("orangered4", "goldenrod1"), 
                    name = NULL, labels = c("Animal food", "Human food")) +
  
  labs(title = "Global soybean production and use share between 1961-2013",
       subtitle = "By types of usage") +
  
  theme(plot.title = element_text(face = 'bold',size = 13),
        plot.subtitle = element_text( size = 12),
        legend.position = "bottom",
        legend.key.size = unit(.5, "cm"),
        legend.text = element_text(size = 10),
        legend.box.spacing = unit(0.2, "cm"),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text.x = element_text( size = 7,face = "bold"),
        axis.text.y = element_text(margin = margin(t = .3, unit = "cm"), size = 7, face = "bold"))

```

#### Figure 2

```{r forest_net_converstion, message = FALSE, fig.alt = "The net loss in global forest area to use land for another purpose between 1990 and 2015"}
forest2 <- forest %>%
  filter(entity == "World")

ggplot(forest2,aes(x=year, y = net_forest_conversion, fill=year)) + 
  geom_area(aes(color = entity), size = 2.5, fill= "lightgray", alpha = 0.5) +
  scale_color_manual(values = c("burlywood4")) +
  scale_y_reverse() +

  
  labs(title = "Global Net Forest Conversion (hectares)", x = "Year", y = NULL) +
  
  theme(legend.position = "none",
        plot.title = element_text(face = 'bold',size = 17),
        axis.title.x = element_text(size = 16),
        axis.text.x = element_text( size = 13,face = "bold"),
        axis.text.y = element_text(margin = margin(t = .3, unit = "cm"), size = 13, face = "bold"))
```

### Discussion

(1-3 paragraphs)
In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots.
Speculate about why the data looks the way it does.


## Presentation

Our presentation can be found [here](presentation/presentation.html).

## Data 

Ritchie, H. and Roser, M. 2021, "Forests and Deforestation". Published online at OurWorldInData.org. Retrieved from: 'https://ourworldindata.org/forests-and-deforestation' on September 13, 2021 [Online Resource]

## References

1. [Brazilian Presidents](https://en.wikipedia.org/wiki/List_of_presidents_of_Brazil)
2. Bush, M. B., Silman, M. R., McMichael, C., & Saatchi, S. (2008). Fire, climate change and biodiversity in Amazonia: a Late-Holocene perspective. Philosophical transactions of the Royal Society of London. Series B, Biological sciences, 363(1498), 1795–1802. https://doi.org/10.1098/rstb.2007.0014
3. [Soybeans in Argentina](https://croplife.org/news-views/sharing-the-story/soybeans-from-argentina/#:~:text=The%20country%20grew%2020%20million,cereal%2C%20automotive%20and%20petrochemical%20exports.&text=The%20health%20of%20the%20soybean,therefore%2C%20of%20great%20national%20interest.)
3. [Our World in Data](https://ourworldindata.org/forests-and-deforestation)
4. [Deforestation](https://ourworldindata.org/deforestation)  
5. [Share of forest area](https://ourworldindata.org/forest-area)  
6. [Drivers of deforestation](https://ourworldindata.org/drivers-of-deforestation)  
7. [Deforestation by commodity](https://ourworldindata.org/grapher/deforestation-by-commodity)  
8. [Soybean production and use](https://ourworldindata.org/soy)  
9. [Palm oil production](https://ourworldindata.org/palm-oil) 
10.[Net Forest Conversion](https://www.clientearth.org/latest/latest-updates/stories/what-is-forest-conversion/)
