---
title: "Project proposal"
author: "Team 5 Look Live"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
```

## Dataset

```{r load_data}
forest <- read.csv("../data/forest.csv")
forest_area <- read.csv("../data/forest_area.csv")
brazil_loss <- read.csv("../data/brazil_loss.csv")
vegetable_oil <- read.csv("../data/vegetable_oil.csv")
soybean_use <- read.csv("../data/soybean_use.csv")
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

We chose the global deforestation data from https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-04-06/readme.md,
which was  sourced from Our World in Data. 

The dataset has 5 elements: `forest`, `forest_change`, `brazil_loss`, `soybean_use`, and `vegetable_oil`. The set `forest` contains the change every 5 years in for forest area in conversion. It comes originally from the UN Food and Agriculture Organization’s Forest Resources Assessment. "Since year-to-year changes in forest cover can be volatile, the UN FAO provide this annual data averaged over five-year periods." `forest_area` contains the change in global forest area as a percent of global forest area. `brazil_loss` contains the loss of Brazilian forest due to specific types of deforestation, such as `fire` and `small_scale_clearing`. `soybean_use` contains the soybean production and use by year and by country. Finally, the set `vegetable_oil` contains vegetable oil production by crop type and year.

We chose this dataset because we were interested to see the effects of soybean production on deforestation in Brazil and globally. The data also had minimal missing values, and multiple datasets, so we were able to brainstorm several ideas we'd be interested in. 

## Questions

Question one: What is the relationship between global soybean use and deforestation in Brazil? (specifically, deforestation in Brazil due to commercial crops? specifically, USA? specifically, which type of use?)

Question two: How does the proportions of soybean production used for human food versus for feeding animal change over time? and whether this change in proportions is related to global deforestation development over time?

## Analysis plan



Plan for Question Two: 
1. group all entries in soybean_use.csv by decades; 
2. For each decade category, add up the values for all entries under "human_food", "animal_feed", and "processed";
3. create new variables "human_prop" and "animal_prop" that represents the proportions of soybean production used for human food and for animal feed out of the totality each decade;
4.plot "human_prop" and "animal_prop" as dots in a graph where the x-axis represents "decade" and the y-axis represents proportion percentage between 0 to 1. Connect all dots under each variable - "human_prop" and "animal_prop" - with lines in differentiating shape and color.
5. plot percent of global forest area, "forest_area", as points in the above graph and connect the dots with a different colored line.
6. Observe and compare the lines.




